### 2020-08-17 paper/sex_selection_spacing-ver4.tex
Editing and update with Demography style guide

\title{Birth Spacing in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

% Demography abstract
\noindent 
In India, the most substantial lengthening in Hindu women's birth intervals over the last 
four decades occurred for the women most likely to use sex-selective abortions. 
For well-educated women with two girls, the median birth interval to the third child 
increased by nearly 15 months---and the 75th percentile interval by 21 months---as the 
sex ratio approached 80 percent boys. 
Because of the rapid lengthening in spacing, standard fertility rates significantly 
overestimated how fast cohort fertility fell. 
Despite convergence, cohort fertility is still ten to twenty percent higher than the 
fertility rate, and above replacement level for all but the best-educated urban women. 
With the lengthening from sex selection, some women show a reversal of the traditional 
spacing pattern; instead of the shortest birth interval in the absence of sons, they now 
have the longest. 
Finally, infant mortality has declined over time for all groups, but the decline was the 
fastest for the less educated. 
Short birth spacing is still associated with higher mortality, although considerably less 
so for the best-educated women. 
There is no evidence that the increasing use of sex selection is associated with higher 
infant mortality. 
Neither does it appear that the use of sex selection is declining.

@@

% Background
India experienced many positive changes over the last four decades:
the economy grew substantially,
schooling increased for males and females,
and the total fertility rate fell to 2.2 
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

However, India also witnessed the advent of sex selection---the selective abortion of 
female fetuses based on prenatal sex determination---with the introduction of ultrasound 
in the mid-eighties.
Combined with continued strong son preference, the result was a dramatic increase in the 
male-to-female ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.

% Questions addressed and why valuable
The main question here is: how did birth spacing respond to these changes, especially 
the spread of sex selection? 
The motivation is twofold.
First, prior research has failed to appreciate that each abortion increases the
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
but we know little about how much. 
Second, more female education, higher household income, and low and declining female labor 
force participation all likely impact birth spacing. 
Hence, the combined changes in birth spacing may outpace what we have observed in 
other countries.

Studying birth spacing contributes to our understanding of fertility decisions, but, equally 
important, birth spacing affects the reliability of our fertility measures and may 
impact mortality. 
I, therefore, address two additional questions.

First, did changes in birth spacing bias the fertility estimates for India? 
With longer spacing, mothers will be older at each parity, and this tempo effect makes the 
total fertility rate a downward biased estimate of cohort fertility
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}. 
Hence, if birth intervals increased substantially, India's fertility may be higher 
than generally accepted.

Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?
In India, birth intervals have traditionally been shorter with fewer sons, contributing 
to the higher mortality risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Hence, longer birth spacing, whether from sex selection or secular changes, may reduce 
mortality \citep{Conde-Agudelo2012,Molitoris2019}.
However, a counteracting effect is possible if longer birth intervals arise from multiple 
abortions since the short duration between \emph{pregnancies} could increase mortality. 
Nevertheless, there has been no attempt to examine how infant mortality changed with 
the increase in sex selection.

@@

The primary outcomes I examine are 25th, 50th, and 75th percentile birth intervals, the 
sex ratio at birth, and the likelihood of giving birth. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key variables are maternal education, the sex composition of previous 
children, and area of residence.

@@

First, birth intervals lengthen over the four decades, more so the higher the parity, the 
more educated the woman, and the higher the percentile. 
Women most likely to use sex selection---well-educated with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some 
women with no sons having \emph{longer} birth intervals than those with sons. 
Those least likely to use sex selection---less-educated in rural areas---still follow the 
traditional pattern of short spacing when they have girls and only some evidence of sex 
selection. 
The likelihood of a very short birth interval changed little.

Second, the fertility rate substantially overestimated how fast cohort fertility fell in the 
1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still ten to 
twenty percent higher than the fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

@@

I divide eduction into four groups: no education, one to seven years, eight to eleven 
years, and twelve and more years.
The latter two correspond to having completed primary and completed secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and eleven and twelve for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and five to sevens years of schooling completed and the eight to nine 
and ten to eleven years of schooling completed.
This grouping allows me to capture the differences across education levels discussed below, 
while also having groups large enough for the empirical method.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
twenty years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90 percent for the 1930s cohorts to 
less than 20 percent for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than sixty percent for the 1990s cohorts, with about half in the eight to eleven 
group and the other half in the twelve plus group.
In urban areas, the proportions with no education or one to seven years have each declined 
to just below ten percent.
Most of the increase in urban female education came from the twelve plus group, which 
now account for more than half of all urban women for the lastest cohorts.

@@

The standard economic argument for shorter spacing with increasing female education 
is that parents incur time costs when they have children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shortening birth spacing to 
take advantage of economies of scale in childrearing \citep{Vijverberg1982}.

@@

The low and declining female labor force participation, especially for younger women, 
suggests that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70 percent higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to own wages \citep{Bhargava2018}.

@@

For birth intervals, I estimate a set of percentile birth intervals. 
I first calculate for each woman when there is a given percent chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80 percent parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60 percent mark on her survival 
curve. 
I then and add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

@@

The predicted sex ratio is the weighted average of individual predicted sex ratios, using 
parity progression probabilities as weights.
To find the individual sex ratio, I estimate the percent of births that are boys at $t$, 
conditional on not having had a child before $t$.
Weighting the percentages boys with the likelihood of exiting the spell with a birth across all 
$t$ gives the predicted percent of boys over the entire spell for an individual.%
\footnote{
Imagine $T=2$. 
If 54 and 66 percent of births are boys and the likelihood of giving birth 20 and 40 percent, 
then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent boys. 
}

@@

I restrict the sample to Hindus for two reasons.
First, Hindus are the majority population group, about 80 percent of India’s population.
Second, the prior literature shows that son preference and use of sex selection vary 
substantially between Hindus and the second largest group, Muslims. 
Combining them and assuming that the baseline hazard is the same would lead to biased 
results.
Because of space constraints and the relatively small number of observations once split 
by education and periods, I do not provide separate results for Muslims or any of the 
remaining groups, such as Sikhs, Jains, and Christians.

@@

To ensure that there are enough births for the method to work, I censor spells at 96 
months (eight years) after a woman can first give birth, equivalent to 105 months 
after the birth of the prior child.
Less than one percent of observed births occur after the cut-off.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

@@

Direct information on the use of sex selection is not available, so I compare periods 
based on the changes in access and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--83, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
% Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Finally, although sex selection increased even after 1994, we may have passed a 
turning point in its use in the mid-2000s \citep{Das_Gupta2009,Kumar2012,Bongaarts2013,Diamond-Smith2015}.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentiles birth intervals in 
months, the sex ratio, and the probability of having a birth for each spell
by education levels and area of residence.%
\footnote{
Results for urban women without education, rural women with twelve or more years of 
education, and the fourth spell for women with twelve or more years of education are
in the online appendix because of relatively small samples.
}
The sex ratio graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.
The underlying values with bootstrapped standard errors are available in the online 
appendix.

\captionsetup[figure]{skip = -16}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_low_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with no education by spell, sex composition, and period}
\label{fig:spacing_low_rural}
\end{figure}

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_low_urban}
% \caption{Percentile birth intervals, sex ratios, and parity progression  
% for urban women with no education by spell, sex composition, and period}
% \label{fig:spacing_low_urban}
% \end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_med_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with one to seven years of education by spell, sex composition, and period}
\label{fig:spacing_med_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_med_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with one to seven years of education by spell, sex composition, and period}
\label{fig:spacing_med_urban}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_high_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with eight to eleven years of education by spell, sex composition, and period}
\label{fig:spacing_high_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_high_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with eight to eleven years of education by spell, sex composition, and period}
\label{fig:spacing_high_urban}
\end{figure}

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_highest_rural}
% \caption{Percentile birth intervals, sex ratios, and parity progression  
% for rural women with twelve or more years of education by spell, sex composition, and period}
% \label{fig:spacing_highest_rural}
% \end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectration=true]{bs_highest_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with twelve or more years of education by spell, sex composition, and period}
\label{fig:spacing_highest_urban}
\end{figure}

@@

% [Sex ratios]
Second, 
the spread of sex selection shows clearly in the sex ratios of next births, which has 
become more male-dominated for women with no sons.
The percent of births that were boys increased faster, the higher the education and the 
higher the parity.
There are no clear trends for the other sex compositions.
Within a given spell and period combination and in the absence of a son, sex ratios are 
higher, the more educated the mother, and the higher the parity. 
Sex ratios are also higher in urban than in rural areas.
Some women with one son also show an unnaturally high percentage of boys, although the 
failing fertility makes these estimates noisy.

@@

% [25th percentile birth intervals]
A remarkably high proportion of birth intervals are still very short.
For all but the most educated, more than 25 percent have their second and third child 
within 24 months of the previous birth, substantially below the 24 months between 
\emph{pregnancies} the WHO recommends.
Furthermore, despite higher parities' more substantial increases in birth intervals, even 
the 25th percentile birth intervals for the fourth spell are around 24 months for women 
with less than eight years of education.

@@

Birth intervals appeared to lengthen the most for women the least likely to work.
For example, from lowest to higher education, the average third spell birth intervals for 
urban women with one boy and one girl increased by 2.7, 3.4, 5.8, and 1.8 months over the 
four decades.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0.
See the online appendix tables for the average birth intervals.
}
Hence, women with the lowest labor force participation, those with eight to eleven years 
of education, also saw the largest increases in average spacing, possibly driven by the 
substantial improvement in household income for this group from economic growth.

% [75th percentile birth intervals]
The most substantial changes occurred in the 75th percentile birth intervals, where the 
more the parity progression probabilities declined, the more the birth interval lengthen. 
For example, the probability of a fourth birth for urban women with eight to eleven years 
of education and two sons and a girl has declined by almost forty percentage points as the 
75th percentile birth interval increased by 22 months.
Compare this with no education, rural women with a boy as their first child, where the 
probability of a third birth declined by less than six percentage points while the birth 
interval increased by only slightly over two months.

@@

A clear illustration of how the combination of son preference and the introduction of 
sex selection affected birth spacing comes from the third spell of the best-educated urban 
women.
With two girls, almost 80 percent of the third births are boys, and the 75th percentile 
birth interval is close to 70 months. 
This interval is about 13 months longer than if they had at least one son and represents 
an increase of almost 21 months over the four decades.

@@

The 75th percentile changes are the most dramatic, but sex selection also impacts the 
25th and median birth intervals. 
For the best-educated urban women with two girls, the 25th percentile birth interval 
increased by six months, or 23 percent, while the median percentile birth interval 
increased by 15 months (43 percent).

Not surprisingly, given these effects of sex selection, the third spell for the 
best-educated women shows the clearest reversal in the spacing pattern; 
the birth intervals with two girls are consistently longer than the intervals with one or 
two boys, no matter the percentile used.
A similar reversal, although more muted, occurred for the third spell for urban women 
with one to seven years of education and both urban and rural women with eight to
eleven years of education.

@@

Because NFHS-1 was after the introduction of sex selection, I cannot calculate a fertility 
rate in precisely the same manner for a period before sex selection was widely available.
Instead, I calculate the fertility rates for women between 15 and 39 years of age five years 
before the survey month, again using the number of births three years before.
This rate is shown as ``1987--88'' in the table.
Given the relatively low number of births to women 40 to 45 years of age, this approach 
provides the best estimate of the fertility rate when sex selection still was not 
widespread.

@@

Finally, even with the declines in the predicted cohort fertility, it is still mostly 
above replacement.
Only for urban women with twelve or more years of education is the predicted cohort
fertility clearly below 2.1.
Even then, cohort fertility is still more than 0.3 children higher than 
the fertility rate estimate of 1.5.
Furthermore, the predicted cohort fertility numbers are likely too low since I use only 
the first four births and births before the imposed 105-month birth interval censoring.

@@

\captionsetup[figure]{skip = 10}

>>>>

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

% Demography abstract
\noindent 
In India, the most substantial lengthening in Hindu women's birth intervals over the last 
four decades occurred for the women most likely to use sex-selective abortions. 
For well-educated women with two girls, the median birth interval to the third child 
increased by nearly 15 months---and the 75th percentile interval by 21 months---as the 
sex ratio approached 80\% boys. 
Because of the rapid lengthening in spacing, standard fertility rates significantly 
overestimated how fast cohort fertility fell. 
Despite convergence, cohort fertility is still 10 to 20\% higher than the 
fertility rate, and above replacement level for all but the best-educated urban women. 
With the lengthening from sex selection, some women show a reversal of the traditional 
spacing pattern; instead of the shortest birth interval in the absence of sons, they now 
have the longest. 
Finally, infant mortality has declined over time for all groups, but the decline was the 
fastest for the less educated. 
Short birth spacing is still associated with higher mortality, although considerably less 
so for the best-educated women. 
There is no evidence that the increasing use of sex selection is associated with higher 
infant mortality. 
Neither does it appear that the use of sex selection is declining.

@@

% Background
India has experienced many positive changes over the past four decades:
the economy has grown substantially,
educational attainment has increased for males and females,
and the total fertility rate has fallen to 2.2 children
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

However, India also witnessed the advent of sex selection---the selective abortion of 
female fetuses based on prenatal sex determination---with the introduction of ultrasound 
in the mid-1980s.
Combined with a strong continued preference for sons, the result was a dramatic increase 
in the male--female ratio at birth
\citep{das_gupta97,Arnold2002,retherford03b,Guilmoto2012,Portner2015b,Jayachandran2017}.

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, prior research has failed to appreciate that each abortion increases the
interval between births by 6 to 12 months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
although we do not know by how much. 
Second, greater educational attainment by women, higher household income, and the low and 
declining female labor force participation all likely influence birth spacing. 
Thus, the combined changes in birth spacing may outpace what we have observed in 
other countries.

Studying birth spacing contributes to our understanding of fertility decisions, but, equally 
important, birth spacing also affects the reliability of our fertility measures and may 
affect mortality. 
Therefore, I address two additional questions.

First, did changes in birth spacing bias the fertility estimates for India? 
With longer spacing, mothers will be older at each parity, and this tempo effect makes the 
total fertility rate a downward biased estimate of cohort fertility
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}. 
Hence, if birth intervals increased substantially, then India's fertility may be higher 
than generally accepted.

Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?
In India, birth intervals have traditionally been shorter with fewer sons, contributing 
to the higher mortality risk for girls
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.
Therefore, longer birth spacing, whether from sex selection or secular changes, may reduce 
mortality \citep{Conde-Agudelo2012,Molitoris2019}.
However, a counteracting effect is possible if longer birth intervals arise from multiple 
abortions because the short duration between pregnancies could increase mortality. 
Nevertheless, there has been no attempt to examine how infant mortality has changed with 
the increase in sex selection.

@@

The primary outcomes I examine are the 25th, 50th, and 75th percentile birth intervals;
the sex ratio at birth; and the likelihood of giving birth. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key variables are maternal education, the sex of previous children, and the area of 
residence.

@@

First, birth intervals lengthen over the four decades---more so the higher the parity, the 
more educated the woman, and the higher the percentile. 
Women most likely to use sex selection---well-educated women with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some 
women with no sons having \emph{longer} birth intervals than those with sons. 
Those least likely to use sex selection---less-educated women in rural areas---still follow the 
traditional pattern of short spacing when they have girls with only some evidence of sex 
selection. 
The likelihood of a very short birth interval changed little.

Second, the fertility rate substantially overestimated how fast cohort fertility fell in the 
1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still 10 to 
20\% higher than the fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

@@

I divide eduction into four groups: no education, one to seven years, 8 to 11 years, and 
12 and more years.
The latter two correspond to having completed primary and completed secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and 11 and 12 for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and five to seven years of schooling completed and the eight to nine 
and 10 to 11 years of schooling completed.
This grouping allows me to capture the differences across education levels discussed below, 
while also having groups large enough for the empirical method.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
twenty years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90\% for the 1930s cohorts to 
less than 20\% for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than 60\% for the 1990s cohorts, with about half in the 8 to 11 group and the 
other half in the 12 plus group.
In urban areas, the proportions with no education or one to seven years have each declined 
to just below 10\%.
Most of the increase in urban female education came from the 12 plus group, which 
now account for more than half of all urban women for the lastest cohorts.

@@

The standard economic argument for shorter spacing with increasing female education 
is that parents incur time costs when they have children \citep{Hotz1997,schultz97}.
Suppose having children requires the mother to reduce her market work. 
Parents can then lower this cost of children by shortening birth spacing to take advantage 
of economies of scale in childrearing  \citep{Vijverberg1982}.

@@

The low and declining female labor force participation, especially for younger women, 
suggests that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to own wages \citep{Bhargava2018}.

@@

For birth intervals, I estimate a set of percentile birth intervals. 
I first calculate for each woman when there is a given percent chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80\% parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60\% mark on her survival 
curve. 
I then add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

@@

The predicted sex ratio is the weighted average of individual predicted sex ratios, using 
parity progression probabilities as weights.
To find the individual sex ratio, I estimate the percent of births that are boys at $t$, 
conditional on not having had a child before $t$.
Weighting the percentages boys with the likelihood of exiting the spell with a birth across all 
$t$ gives the predicted percent of boys over the entire spell for an individual.%
\footnote{
Imagine $T=2$. 
If 54 and 66\% of births are boys and the likelihood of giving birth 20 and 40\%, 
then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$\% boys. 
}

@@

I restrict the sample to Hindus for two reasons.
First, Hindus are the majority population group, about 80\% of India’s population.
Second, the prior literature shows that son preference and use of sex selection vary 
substantially between Hindus and the second largest group, Muslims. 
Combining them and assuming that the baseline hazard is the same would lead to biased 
results.
Because of space constraints and the relatively small number of observations once split 
by education and periods, I do not provide separate results for Muslims or any of the 
remaining groups, such as Sikhs, Jains, and Christians.

@@

To ensure that there are enough births for the method to work, I censor spells at 96 
months (eight years) after a woman can first give birth, equivalent to 105 months 
after the birth of the prior child.
Less than 1\% of observed births occur after the cut-off.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

@@

Direct information on the use of sex selection is not available, so I compare periods 
based on the changes in access and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971.
Reports of sex determination appeared around 1982--1983, and the number of clinics 
quickly increased \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Prenatal Diagnostic Techniques Act made determining and communicating 
the sex of a fetus illegal.%
\footnote{
% Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
There is little evidence that the ban significantly affected sex ratios \citep{Das-Gupta2016}.
}
Finally, although sex selection increased even after 1994, we may have passed a 
turning point in its use in the mid-2000s \citep{Das_Gupta2009,Kumar2012,Bongaarts2013,Diamond-Smith2015}.

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentiles birth intervals in 
months, the sex ratio, and the probability of having a birth for each spell
by education levels and area of residence.%
\footnote{
Results for urban women without education, rural women with twelve or more years of 
education, and the fourth spell for women with twelve or more years of education are
in the online appendix because of relatively small samples.
}
The sex ratio graphs also show the natural sex ratio, approximately 51.2\% boys
\citep{ben-porath76b,jacobsen99,Portner2015b}.
The underlying values with bootstrapped standard errors are available in the online 
appendix.

\captionsetup[figure]{skip = -16pt}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with no education by spell, sex composition, and period}
\label{fig:spacing_low_rural}
\end{figure}

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_urban}
% \caption{Percentile birth intervals, sex ratios, and parity progression  
% for urban women with no education by spell, sex composition, and period}
% \label{fig:spacing_low_urban}
% \end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with one to seven years of education by spell, sex composition, and period}
\label{fig:spacing_med_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with one to seven years of education by spell, sex composition, and period}
\label{fig:spacing_med_urban}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_high_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with eight to eleven years of education by spell, sex composition, and period}
\label{fig:spacing_high_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_high_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with eight to eleven years of education by spell, sex composition, and period}
\label{fig:spacing_high_urban}
\end{figure}

% \begin{figure}
% \centering
% \includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_highest_rural}
% \caption{Percentile birth intervals, sex ratios, and parity progression  
% for rural women with twelve or more years of education by spell, sex composition, and period}
% \label{fig:spacing_highest_rural}
% \end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_highest_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with twelve or more years of education by spell, sex composition, and period}
\label{fig:spacing_highest_urban}
\end{figure}

@@

% [Sex ratios]
Second, 
the spread of sex selection shows clearly in the sex ratios of next births, which has 
become more male-dominated for women with no sons.
The percentage of births that were boys increased faster, the higher the education and the 
higher the parity.
There are no clear trends for the other sex compositions.
Within a given spell and period combination and in the absence of a son, sex ratios are 
higher, the more educated the mother, and the higher the parity. 
Sex ratios are also higher in urban than in rural areas.
Some women with one son also show an unnaturally high percentage of boys, although the 
failing fertility makes these estimates noisy.

@@

% [25th percentile birth intervals]
A remarkably high proportion of birth intervals are still very short.
For all but the most educated, 25\% or more have their second and third child 
within 24 months of the previous birth.
These intervals are substantially below the 24 months between 
\emph{pregnancies} the WHO recommends.
Furthermore, despite higher parities' more substantial increases in birth intervals, even 
the 25th percentile birth intervals for the fourth spell are around 24 months for women 
with less than eight years of education.

@@

Birth intervals appeared to lengthen the most for women the least likely to work.
For example, from lowest to higher education, the average third spell birth intervals for 
urban women with one boy and one girl increased by 2.7, 3.4, 5.8, and 1.8 months over the 
four decades.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0.
See the online appendix tables for the average birth intervals.
}
Hence, women with the lowest labor force participation, those with 8 to 11 years 
of education, also saw the largest increases in average spacing, possibly driven by the 
substantial improvement in household income for this group from economic growth.

% [75th percentile birth intervals]
The most substantial changes occurred in the 75th percentile birth intervals, where the 
more the parity progression probabilities declined, the more the birth interval lengthen. 
For example, the probability of a fourth birth for urban women with eight to eleven years 
of education and two sons and a girl has declined by almost 40 percentage points as the 
75th percentile birth interval increased by 22 months.
Compare this with no education, rural women with a boy as their first child, where the 
probability of a third birth declined by less than six percentage points while the birth 
interval increased by only slightly over two months.

@@

A clear illustration of how the combination of son preference and the introduction of 
sex selection affected birth spacing comes from the third spell of the best-educated urban 
women.
With two girls, almost 80\% of the third births are boys, and the 75th percentile 
birth interval is close to 70 months. 
This interval is about 13 months longer than if they had at least one son and represents 
an increase of almost 21 months over the four decades.

@@

The 75th percentile changes are the most dramatic, but sex selection also impacts the 
25th and median birth intervals. 
For the best-educated urban women with two girls, the 25th percentile birth interval 
increased by six months, or 23\%, while the median percentile birth interval 
increased by 15 months (43 percent).

Not surprisingly, given these effects of sex selection, the third spell for the 
best-educated women shows the clearest reversal in the spacing pattern; 
the birth intervals with two girls are consistently longer than the intervals with one or 
two boys, no matter the percentile used.
A similar reversal, although more muted, occurred for the third spell for urban women 
with one to seven years of education and both urban and rural women with 8 to 11 years of 
education.

@@

Because NFHS-1 was after the introduction of sex selection, I cannot calculate a fertility 
rate in precisely the same manner for a period before sex selection was widely available.
Instead, I calculate the fertility rates for women between 15 and 39 years of age five years 
before the survey month, again using the number of births three years before.
This rate is shown as ``1987--1988'' in the table.
Given the relatively low number of births to women 40 to 45 years of age, this approach 
provides the best estimate of the fertility rate when sex selection still was not 
widespread.

@@

Finally, even with the declines in the predicted cohort fertility, it is still mostly 
above replacement.
Only for urban women with 12 or more years of education is the predicted cohort
fertility clearly below 2.1 children.
Even then, cohort fertility is still more than 0.3 children higher than 
the fertility rate estimate of 1.5.
Furthermore, the predicted cohort fertility numbers are likely too low since I use only 
the first four births and births before the imposed 105-month birth interval censoring.

@@

\captionsetup[figure]{skip = 10pt}
